Association between trauma severity models and opportunities for improvement:

### 2022-10-25 manuscript.Rmd

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response <!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity models
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity models. Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

@@

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it doesn't account for situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].

@@

The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences. M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring. However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether trauma severity models have better performance when applied to the entire population of trauma patients.

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

The trauma registry at KUH is also submitted to SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines. The inclusion criteria are given below. All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma clinic. Data is collected from the patient's electronic medical records. The database contains data on patients from 2012 to 2021.

SweTrau follows the Utstein Trauma Template for reporting of data following major trauma. The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment. Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS at discharge, highest level of care, and transfers to other units.

@@

### Study-specific exclusion criteria
Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity. Patients that were dead on arrival were also excluded. In addition, this is a complete case study and all patients that lacked parameters necessary to carry out the analysis were excluded.

@@

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
### Outcome
The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH. OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".

@@

All patient information used in this study is anonymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request. <!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->

>>>>

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015]<!-- Martin cite -->, is one of the leading causes of death worldwide, constituting nearly 8% of global deaths annually [@whoepi]<!-- Cite the most recent Global Burden of Disease study instead-->. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

## Trauma severity models
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity models. Trauma severity models attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. Trauma severity models were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of trauma severity models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps)<!-- Consider not abbreviating Ps, as it's quite short anyway and is a "non-standard" abbreviation-->.

@@

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it doesn't<!-- Don't use contractions --> account for situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].

@@

The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences. M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths<!-- I would not include preventable deaths here-->. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The World Health Organisation (WHO) themselves note in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

Another potential audit filter for M&M review, that is already used at some clinics [@whoguidelines], is trauma severity scoring. However, the extent to which trauma severity models are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution<!-- You haven't mentioned "this institution" before, so it's hard to understand what you refer to. I suggest that state the actual centre--> found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether trauma severity models have better performance when applied to the entire population of trauma patients.

@@

This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH)<!-- Use the full hospital name-->: the trauma registry and the trauma care quality database. The trauma care quality database is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common trauma severity models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

The trauma registry at KUH is part of SweTrau, Sweden's national trauma registry, and thereby follows SweTrau guidelines. The inclusion criteria are given below. All trauma patients meeting the inclusion criteria are registered in the trauma registry by nurses at the trauma centre. Data is collected from the patient's electronic medical records. The database<!-- Trauma registry?--> contains data on patients from 2012 to 2021.

SweTrau follows the Utstein Trauma Template for reporting of data following major trauma. The registered data includes prehospital vital signs and management, vital signs and management on arrival at hospital, information about the type of injury, and the primary method of treatment. Follow-up variables include, but are not limited to, survival at 30 days, days in ventilator, GCS<!-- Do you mean GOS?--> at discharge, highest level of care, and transfers to other units.

@@

### Study-specific exclusion criteria
Patients <15 years of age were excluded as paediatric trauma is a separate clinical entity<!-- And are not part of the same review process-->. Patients that were dead on arrival were also excluded. In addition, this is a complete case study and all patients that lacked parameters necessary to carry out the analysis were excluded.

@@

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

### Outcome
The studied outcome is the binary variable "opportunities for improvement" (OFI), as identified by the multidisciplinary M&M conference at KUH. OFI is coded as either "Yes - at least one opportunity for improvement identified" or "No - no opportunities for improvement identified".

@@

All patient information used in this study is pseudononymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request. <!-- https://meetings.lg.se/pdfFile/pdfFile.aspx?objectID=%2F7Iqgtt89nW2zhRjMSNZ%2Fw%3D%3D -->
